---
title: "week06_interactive_analysis_version2_collins"
author: "Mallory Collins"
date: "2026-04-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r load_data}
if (!require("MASS")) install.packages("MASS")
if (!require("plotly")) install.packages("plotly")
if (!require("crosstalk")) install.packages("crosstalk")


library(tidyverse)
library(MASS)
library(plotly)
library(crosstalk)

data("birthwt")

data <- birthwt %>%
  as_tibble() %>%
  mutate(
    low = factor(low, levels = c(0,1), labels = c("not_low","low")),
    smoke = factor(smoke, levels = c(0,1), labels = c("no","yes")),
    ht = factor(ht, levels = c(0,1), labels = c("no","yes")),
    ui = factor(ui, levels = c(0,1), labels = c("no","yes")),
    race = factor(race, levels = c(1,2,3), labels = c("white","black","other"))
  )
```


```{r inspect_data}
# Basic structure
glimpse(data)

# Dimensions
dim(data)

# Variable names
names(data)

# Create a clean summary table
summary_table <- data %>%
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    mean_weight = mean(lwt, na.rm = TRUE),
    mean_birth_weight = mean(bwt, na.rm = TRUE)
  )

# Display nicely formatted table
knitr::kable(
  summary_table,
  caption = "Summary statistics for birth weight dataset",
  digits = 2
)
```
What does each row represent?
variables
What is the response variable?
birth weight
Which variables are categorical vs numeric?
Categorical: Low, race, smoking, ptl, ht
numerical: ui, ftv, bwt


```{r box_plot}
smoke_plot <- ggplot(data, aes(x = smoke, fill = low)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("not_low" = "#7FB77E", "low" = "red")) +
  labs(
    title = "Proportion of Low Birth Weight by Smoking Status",
    y = "Proportion"
  ) +
  theme_minimal()

smoke_plot
```

```{r static_plot}
weight_plot <- ggplot(data, aes(x = low, y = lwt, fill = low)) +
  geom_boxplot() +
  labs(
    title = "Mother's Weight by Birth Weight Category",
    x = "Birth Weight Category",
    y = "Mother's Weight"
  ) +
  theme_minimal()

weight_plot
```
Do smokers appear more likely to have low birth weight babies?
yes
Does maternal weight differ between groups?
no
What patterns stand out?
There is not a large differentce between the percent of low birth weight babies between smoking and non smoking mothers from the bar chart

```{r shared_data}
shared_data <- SharedData$new(data)

bscols(
  widths = c(3, 9),
  list(
    filter_select(
      id = "smoke_filter",
      label = "Smoking status",
      sharedData = shared_data,
      group = ~smoke
    ),
    filter_slider(
      id = "age_filter",
      label = "Mother's age",
      sharedData = shared_data,
      column = ~age
    )
  ),
  plot_ly(
    shared_data,
    x = ~lwt,
    y = ~bwt,
    color = ~low,
    type = "scatter",
    mode = "markers"
  )
)
```
What happens when you change the filters?
you can isolate different combinations of data depending on the mothers smoking status and age.
Why might interactive plots be useful?
You can compare data without making several graphs

Why is a linear model not appropriate here?
because we are comparing the differences between two sets of data
What type of response variable do we have?
binary

```{r logistic_regression}
# Fit model
model1 <- glm(low ~ smoke, data = data, family = binomial)

# Create predicted values
data$pred_prob <- predict(model1, type = "response")

# Plot
ggplot(data, aes(x = smoke, y = pred_prob, color = smoke)) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", size = 4, color = "black") +
  labs(
    title = "Predicted Probability of Low Birth Weight by Smoking Status",
    y = "Predicted Probability",
    x = "Smoking Status"
  ) +
  theme_minimal()
```

```{r expand_model}
model2 <- glm(low ~ smoke + age + lwt + ht + ui, data = data, family = binomial)
summary(model2)
```

Is smoking a significant predictor?
No becuase it has a p value above 0.05
What direction is the effect?
positive
Do additional variables improve the model?
Additional variables like age has a more significant association with low birth weight than smoking
What does this mean biologically?
Smoking has some effect on birth weight but it is not a significant factor that contributes to low birth weight. Low birth weight is probably a combination of multiple variables that can increase the likelyhood of low birth weight. 


The smoking of the mother does not have a significant effect on low birthweight. This result is seen by a p value greater than 0.05 by sorting binary data, if a newborn had low birthweught or not and if the mother smokes. Therefore we accept the null hypothesis that smking does not have a significatn effect on birth weight.

What biological or health factors might explain these results?
smoking could also lead to hypeertension and low weight of the mother, other variables which had a lower p value than smoking alone.
What are limitations of this dataset?
Geographical regions, income, and acess to healcare are also factors not reported that may have an effect on birth weight.
How does this compare to your expectations?
I was surprised that smoking alone is not associated significantly with low birth weight, but it makes sense that low birthweight is the outcome of an accumulation of variables.


